---
title: "Class 12 HW: Q13 and Q14"
format: pdf
toc: true
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
table(expr$geno)
```
This result shows the sample size for each genotype.

```{r}
tapply(expr$exp, expr$geno, median)
```

This result shows the corresponding median expression level for each genotype.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```

From the boxplot, relative expression value between A/A and G/G is clearly different - order of expression level is genotype A/A > A/G > G/G. Thus, the SNP does affect the expression level of ORMDL3.
