---
title: "Class 5: Data Viz with ggplot2"
author: "Seunghoon Oh (A19372132)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R"* graphics like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:
```{r}
plot(cars)
```


Before I can use ggplot2 I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our quarto doc (we run it once only in our R console) as it would re-install every time we render our quarto report.

Once installed we need to load up the package into our R brain (session).

```{r}
library(ggplot2)
```


The main function in **ggplot2** package is called `ggplot()`.

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth() +
  theme_bw () +
  labs(title="Stopping distance of old cars", x="Speed (MPH)", y="Distance (ft)")
```

Q. Can you make the `geom_smooth()` function produce straight line fit to the data and turn off the "gray" error?

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  theme_bw () +
  labs(title="Stopping distance of old cars", x="Speed (MPH)", y="Distance (ft)")
```

---

## Gene expression figure

Import the data to plot

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes, 10)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  labs(title="Gene Expression Changes Upon Drug Treatment", subtitle = "Just another scatter plot made with ggplot", x= "Control (no drug)", y= "Drug Treatment") +
  scale_color_manual(values=c("blue", "gray", "red")) +
  theme_bw() +
  theme(legend.position = "none")
```

## Going further

Here we read the famous gapminder dataset:

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many countries are in this dataset?

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset
Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p<-ggplot(gapminder) +
  aes (gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.5)
```

```{r}
p
```

I can add more layers to `p`

```{r}
p +
  facet_wrap(~gapminder$continent)
```

Make a plot for 1977 and 2007 only (not all the years in dataset).

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the year 2007.

```{r, message=FALSE}
library(dplyr)
```

```{r}
g07<-filter(gapminder, year==2007)
g77<- filter(gapminder, year==1977)
g<- filter(gapminder, year == 2007 | year == 1977)

ggplot(g07) +
   aes (gdpPercap, lifeExp, col=continent, size=pop) +
   geom_point(alpha=0.5)

ggplot(g) +
   aes (gdpPercap, lifeExp, col=continent, size=pop) +
   geom_point(alpha=0.5)+
   facet_wrap(~year)
```
> Q. Make a histogram of lifeExp colored by continent (use `fill=continent` or `col=continent`)

```{r, message=FALSE}
ggplot(gapminder) +
  aes(lifeExp, fill=continent) +
  geom_histogram()
```

```{r, message=FALSE}
ggplot(gapminder) +
  aes(lifeExp, col=continent) +
  geom_histogram()
```

> Q. Make a histogram of lifeExp faceted by continent

```{r, message=FALSE}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```


